Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-24 drafts/hybrid_zone.tex
yb biologized maths

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \citep{Kruuk1999,barton2000epistasis}.

As in \citet{sedghifar2015spatial}, we consider where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to a spatial position $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same; 
	this occurs if there is no recombination in this block, or if all lineages generated by recombination events 
	in this block trace back to the $A$ side. 
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
Because the identity of the selected allele determines how selection acts on the haplotype,
and because linked alleles can only move between the $A$ and the $B$ background in heterozygotes, 
a key factor in these models is the density and fecundity of heterozygotes.

@@

In this section we aim to give both heuristics for how lineages move, 
that are helpful in providing intuition and for establishing order-of-magnitude estimates,
and analytics, mostly in terms of partial differential equations needing numerical solution. 
\alisa{For simplicity, we ignore coalescence, which equates to ignoring the effects of drift and pedigree.}
\plr{(say why and what that means: e.g., deterministic dynamics)}

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model.}
\plr{(I presume it's been described, but I don't think we found a reference that actually did this, right?  So maybe say it's ``well-known in the field'' or something?  Also, no-one seems to have looked at lineages, and as our final point says, we actually get some more information out of this perspective.)} \alisa{So, when I sati's been previously described, I'm thinking about the traditional cline theory for a single locus (e.g. Slatkin, May), which describes the time-forward diffusion.}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\Delta p$ is the Laplacian\alisa{, or second derivative.}
The equation is approximate because it 
omits terms of order $s^2$
(so deviations from the prediction occur over $1/s^2$ generations),
and relies on the Central Limit Theorem to approximate generation-to-generation dispersal by the Gaussian
(so if dispersal is non-Gaussian, will fit best over longer time scales, say, tens of generations).  \revpoint{AE}{2}
This says that the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$)
is equal to the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$)
and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
due to local proportions of heterozygotes,
which is $s(2p-1)$ (up to an error of size $s^2$).
This can be derived as in \citet{Nagylaki1975};
for discussion of diploidy see \citet{christiansen1995genotypic}.
As we show below the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
(This sort of rescaling is common in the literature, e.g., \citep{fisher1950frequencies,Slatkin1973}.)  \linelabel{ll:rescaling}
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele at position $x$ moves as Brownian motion driven by dispersal
with mean displacement proportional to the derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(the analogous physical system is a particle that moves at speed $\sigma^2$ in the potential $-\log p$). 
\alisa{What does potential here mean?}
\plr{(It means like a gravity well.  Clear now that it says ``particle''?)}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=-r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $-\E[R p(x-R)]/\E[p(x-R)]$.  \revpoint{1}{9}
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ 
\alisa{(perhaps a brief explanation of how this works; in particular I wasn't sure why were were taking the ratio of two expectations...)}
\plr{(That's what the rest of the sentence is? Work it out, and if you still think it needs more detail, add it?)}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

Suppose we sample an allele today $r$ Morgans from the selected site. 
We will follow its lineage back through time, using $\tau$ to denote ``generations ago''
and reserving $t$ for time measured in the usual direction.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(as described above for a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (from $B$ to $A$),
by the assumption of locally random assortment of alleles
(see Appendix \ref{apx:lineage_derivation} for a more precise description).

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages,
the tendency of a lineage to inherit from regions where its type is more common,
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
Equivalently,
the first term ($\Delta q$) reflects the smoothing action of local dispersal,
the second is the net flux induced by reduced fitness in the hybrid zone,
and the third is the contribution of recombination. 
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\paragraph{Numerical computation}
To determine how the cline at a linked locus is expected to relax, 
\alisa{and to confirm the behavior of our model to previously established results,}
\plr{(got a reference for that?)}
we can solve the partial differential equations \eqref{eqn:cline_pde} and \eqref{eqn:q_pde} numerically.
For instance, Figure \ref{fig:linked_cline_heatmaps} shows a heatmap of $p(x,t,r)$, 
the expected frequency of ancestry $A$ at location $x$ and time $t$ 
at a site at recombination distance $r$ from the selected site,
which is computed as the frequency of ancestry $A$ on each selected background weighted by the freequencies of each background:
$q(x,t,r) = p(x,t) q_A(x,t,r) + (1-p(x,t)) q_B(x,t,r)$. 
The equations are solved numerically in R, using the ReacTran package \citep{soetaert2012reactive}.

>>>>

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \citep{Kruuk1999,barton2000epistasis}, \yb{because, selection on an allele depends on its marginal effect on fitness.  
Exceptionally strong epistatic selection which acts more quickly than free recombination may therefore, yield results that differ slightly from ours. 
Likewise, while we do not model ecological selection, previous theory shows that clines formed by under dominant selection and extransic selection are essential indistinguishable, especially over short time scales when there is little time for clines to wobble  \citep{Barton1979a}}.

As in \citet{sedghifar2015spatial}, we consider where the ancestors of modern-day 
fall across geography as we look further back towards the time of first contact.  
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$,
i.e., if its lineage traces back to a spatial position $x<0$  at the time of secondary contact.
A block of genome is of ancestry $A$ if every locus in it does the same; 
	this occurs if there is no recombination in this block, or if all lineages generated by recombination events 
	in this block trace back to the $A$ side. 
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
Because the identity of the selected allele determines how selection acts on the haplotype,
and because linked alleles can only move between the $A$ and the $B$ background in heterozygotes, 
a key factor in these models is the density and fecundity of heterozygotes.

@@

In this section we aim to give both heuristics for how lineages move, 
that are helpful in providing intuition and for establishing order-of-magnitude estimates,
and analytics, mostly in terms of partial differential equations needing numerical solution. 
\yb{Because of challenges in modeling spatial coalescence, we ignore the effects of drift and the pedigree. 
Therefore, while our model incorporates stochastic migration, the genetics displays deterministic dynamics.}
%\plr{(say why and what that means: e.g., deterministic dynamics)}

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model.}
\plr{(I presume it's been described, but I don't think we found a reference that actually did this, right?  So maybe say it's ``well-known in the field'' or something?  Also, no-one seems to have looked at lineages, and as our final point says, we actually get some more information out of this perspective.)} \alisa{So, when I sati's been previously described, I'm thinking about the traditional cline theory for a single locus (e.g. Slatkin, May), which describes the time-forward diffusion.}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p +  2 p (1-p)  \times s (\frac{p}{2} - 1), %YB Rearranged this equation to allow for cleaner description
%    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
\yb{where $ \frac{d}{dt} p$ is the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$), and  $\Delta p$ is the Laplacian\alisa{, i.e. the second derivative of allele frequency over space ($x$), at time $t$.}
The first term,  $\sigma^2/2 \Delta p$, describes the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$). 
The second term,  $2 p (1-p)  s (p/2-1)$, describes the impact of selection on local allele frequency as the product in local genotypic variance, $2 p (1-p)$, and selection on allele $A$, $s (p/2-1)$. The term $s (p/2-1)$ clearly describes  underdominance --- $A$ is favored when common ($p>.5$) and disfavored when rare ($p<.5$).}      
%YB: Pardon the extreme changes here, I tried to make things more obvious to a biologist
%and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
%due to local proportions of heterozygotes,
%which is $s(2p-1)$ (up to an error of size $s^2$).
%This can be derived as in \citet{Nagylaki1975};
%for discussion of diploidy see \citet{christiansen1995genotypic}.

The equation is approximate because it 
omits terms of order $s^2$
(so deviations from the prediction occur over $1/s^2$ generations),
and relies on the Central Limit Theorem to approximate generation-to-generation dispersal by the Gaussian
(so if dispersal is non-Gaussian, will fit best over longer time scales, say, tens of generations).  \revpoint{AE}{2} 
As we show below, the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
(This sort of rescaling is common in the literature, e.g., \citep{fisher1950frequencies,Slatkin1973}.)  \linelabel{ll:rescaling}
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele at position $x$ moves as Brownian motion driven by dispersal
with mean displacement proportional to the derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common i.e. the left side of the range.   
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(the analogous physical system is a particle that moves at speed $\sigma^2$ in the potential $-\log p$). 
%\alisa{What does potential here mean?}
%\plr{(It means like a gravity well.  Clear now that it says ``particle''?)}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=-r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $-\E[R p(x-R)]/\E[p(x-R)]$.  \revpoint{1}{9}
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ 
%\alisa{(perhaps a brief explanation of how this works; in particular I wasn't sure why were were taking the ratio of two expectations...)}
%\plr{(That's what the rest of the sentence is? Work it out, and if you still think it needs more detail, add it?)}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

\yb{Suppose we sample an allele today $r$ Morgans from the selected site, and  
follow its lineage back through time, using $\tau$ to denote ``generations ago'' (r
reserving $t$ for time measured in the usual direction).
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(as described for a selected allele in equation \eqref{eqn:cline_pde}).
Following this lineage back in time, the identity of the selected allele that ancestor carried at time $\tau$, $Z_\tau$, 
 jumps between $A$ and $B$ with recombination events between the neutral and selected locus (with frequency $r$), that result in a change in the  selected background.    
Thus, by the assumption of locally random assortment of alleles, $Z$ shifts from $A$ to $B$  at rate $r (1-p)$, and $B$ to $A$ at rate $r p$ (see Appendix \ref{apx:lineage_derivation} for a more precise description).}

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages (i.e. the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common (i.e. the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\paragraph{Numerical computation}
To determine how the cline at a linked locus is expected to relax, 
%\alisa{and to confirm the behavior of our model to previously established results,}
%\plr{(got a reference for that?)} 
we can solve the partial differential equations \eqref{eqn:cline_pde} and \eqref{eqn:q_pde} numerically.
For instance, Figure \ref{fig:linked_cline_heatmaps} shows a heatmap of $p(x,t,r)$, 
the expected frequency of ancestry $A$ at location $x$ and time $t$ 
at a site at recombination distance $r$ from the selected site,
which is computed as the frequency of ancestry $A$ on each selected background weighted by the frequencies of each background:
$q(x,t,r) = p(x,t) q_A(x,t,r) + (1-p(x,t)) q_B(x,t,r)$. 
The equations are solved numerically in R, using the ReacTran package \citep{soetaert2012reactive}.
